MG7: Configurable and scalable 16S metagenomics data analysis -- new methods optimized for massive cloud computing

### 2015-09-14 docs/paper/out/paper.tex
add : to def lists -- paper

No abstract yet. Will be here.

@@

\begin{enumerate}
\def\labelenumi{\roman{enumi}.}
\tightlist
\item
  \textbf{Sequence data} the reads are larger, the sequencing depth and
  the number of samples of each project are considerably bigger. The
  first metagenomics studies were very local projects, while nowadays
  the most fruitful studies are done at a global level (international,
  continental, national). This kind of global studies has yielded the
  discovery of clinical biomarkers for diseases of the importance of
  cancer, obesity or inflammatory bowel diseases and has allowed
  exploring the biodiversity of varied earth environments.
\item
  \textbf{The genomics explosion} its effect being felt in this case in
  the reference sequences. The immense amount of sequences available in
  public repositories demands new strategies for curation, update and
  storage of metagenomics reference databases: current models will
  (already) have problems to face the future avalanche of metagenomic
  sequence data.
\item
  \textbf{Cloud computing} the appearance of new models for massive
  computation and storage such as the so-called cloud, or the widespread
  adoption of programming methodologies like functional programming, or,
  more speculatively, dependently typed programming. The immense new
  possibilities that these advances offer must have a direct impact in
  metagenomics data analysis.
\item
  \textbf{Open science} the new social manner to do science,
  particularly so in the case of genomics, brings its own set of
  requirements. Metagenomics evolves in a social and global scenario
  following a science democratization trend in which many small research
  groups from distant countries share a common big metagenomics project;
  this global cooperation demands systems allowing for reproducible data
  analysis, data interoperability, and tools and practices for
  asynchronous collaboration between different groups.
\end{enumerate}

>>>>

The exponential growth of metagenomics is adding a significant plus of
complexity to the big data problem in genomics. In this new scenario
impacted by the wide scale and scope of the projects and by the
explosion of sequence data to be analyzed is especially opportune the
use of new possibilities that cloud computing approaches, new functional
and dependently typed programming languages and new database paradigms
as graph databases offer. To tackle the challenges of big data analysis
in this work we have used these new means to design and develop a new
open source methodology for analyzing metagenomics data, MG7. It
exploits the new possibilities that cloud computing offers to get a
system robust, programmatically configurable, modular, distributed,
flexible, scalable and traceable in which the biological databases of
reference sequences can be easily updated and / or frequently
substituted by new ones or by databases specifically designed for
focused projects. MG7 uses parallelization and distributed analysis
based on AWS, with on-demand infrastructure as the basic paradigm and
allow the definition of complex workflows using a composable system for
scaling/parallelizing stateless computations especially designed for
Amazon Web Services (AWS) that counts with a static reproducible
specification of dependencies and behavior of the different components.
The modelling of the taxonomy tree is done using the new paradigm of
graph databases of Bio4j that facilitates the taxonomic assignment tasks
and the calculation of the taxa abundance values considering the
hierarchic structure of taxonomy tree. MG7 includes the new 16S database
16S-DB7 built with a flexible and sustainable system of updating and
project-driven personalization.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Sequence data:} the reads are larger, the sequencing depth and
  the number of samples of each project are considerably bigger. The
  first metagenomics studies were very local projects, while nowadays
  the most fruitful studies are done at a global level (international,
  continental, national). This kind of global studies has yielded the
  discovery of clinical biomarkers for diseases of the importance of
  cancer, obesity or inflammatory bowel diseases and has allowed
  exploring the biodiversity of varied earth environments.
\item
  \textbf{The genomics explosion:} its effect being felt in this case in
  the reference sequences. The immense amount of sequences available in
  public repositories demands new strategies for curation, update and
  storage of metagenomics reference databases: current models will
  (already) have problems to face the future avalanche of metagenomic
  sequence data.
\item
  \textbf{Cloud computing:} the appearance of new models for massive
  computation and storage such as the so-called cloud, or the widespread
  adoption of programming methodologies like functional programming, or,
  more speculatively, dependently typed programming. The immense new
  possibilities that these advances offer must have a direct impact in
  metagenomics data analysis.
\item
  \textbf{Open science:} the new social manner to do science,
  particularly so in genomics, brings its own set of requirements.
  Metagenomics evolves in a social and global scenario following a
  science democratization trend in which many small research groups from
  distant countries share a common big metagenomics project; this global
  cooperation demands systems allowing for reproducible data analysis,
  data interoperability, and tools and practices for asynchronous
  collaboration between different groups.
\end{enumerate}
